Type-Safe Modular Parsing

### 2017-04-27 SLE_17/sections/CaseStudy.tex
Some changes

\paragraph{Extracting Language Components}
Using the pattern demonstrated in Section~\ref{subsec:language-component}, we extract
reusable components from all the calculi. Each
component, which may contain several syntactical structures,
represents a certain feature of the language. For
example, the \lstinline{VarApp} component below represents variables and
function applications, shown in Figure~\ref{fig:casestudy-varapp}.

\begin{figure}[ht]
\lstinputlisting[linerange=13-29]{code/src/papercode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_VARAPP
\caption{The \inlinecode{VarApp} component in case study.}\label{fig:casestudy-varapp}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{resources/depGraph.pdf}
    \caption{Dependency graph of all calculi and components in case study.}
    \label{fig:dependency}
\end{figure}

As shown in the graph, some components such as \lstinline{VarApp} are
created from scratch, while others such as \lstinline{Typed} are
extended from existing components. Since calculi and components have the similar signature, each calculus
can also be extended and reused directly. For example, calculus \lstinline{FullRef} extends from
calculus \lstinline{FullSimple}.

From the dependency graph, we know
that common components such as \lstinline{VarApp} are reused in
lots of calculi. Such reuse could shorten the code considerably. We
will show this advantage and exam the possible performance penalty in
a quantitative way in the next subsection.

@@

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \multicolumn{3}{ c| }{\bfseries SLOC} & \multicolumn{3}{ c| }{\bfseries Time (ms)} \\ \cline{2-7}
        \multicolumn{1}{|c|}{} & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% \\
      \hline
      \input{resources/CaseStudyTable.tex}
      \hline
      \multicolumn{7}{c}{}
    \end{tabular}
    \caption{Comparison of SLOC and execution time.}
    \label{tab:comparison}
\end{table}

@@

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \ilyaimpl{} & \multicolumn{2}{ c| }{\ourimpl{}} & \multicolumn{2}{ c| }{\ilyalongest{}} & \multicolumn{2}{ c| }{\ourclass{}} \\ \cline{2-8}
        \multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{\bfseries Time} & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% \\
      \hline
        Arith & 741 & 913 & +23.2 & 793 & +7.0 & 932 & +25.8 \\
        Untyped & 770 & 1018 & +32.2 & 821 & +6.6 & 1007 & +30.8 \\
        FullUntyped & 1297 & 1854 & +42.9 & 1343 & +3.5 & 1767 & +36.2 \\
        TyArith & 746 & 888 & +19.0 & 772 & +3.5 & 918 & +23.1 \\
        SimpleBool & 1376 & 1782 & +29.5 & 1494 & +8.6 & 1824 & +32.6 \\
        FullSimple & 1441 & 2270 & +57.5 & 1574 & +9.2 & 2226 & +54.5 \\
        Bot & 1080 & 1287 & +19.2 & 1078 & -0.2 & 1306 & +20.9 \\
      %\hline
        %\multicolumn{1}{|c|}{\dots} & \multicolumn{7}{c|}{\dots} \\
      %\hline
        FullRef & 1438 & 2291 & +59.3 & 1544 & +7.4 & 2142 & +49.0 \\
        FullError & 1410 & 1946 & +38.0 & 1524 & +8.1 & 1981 & +40.5 \\
        RcdSubBot & 1247 & 1524 & +22.2 & 1285 & +3.0 & 1612 & +29.3 \\
        FullSub & 1320 & 1979 & +49.9 & 1393 & +5.5 & 1899 & +43.9 \\
        FullEquiRec & 1407 & 2200 & +56.4 & 1561 & +10.9 & 2156 & +53.2 \\
        FullIsoRec & 1492 & 2253 & +51.0 & 1648 & +10.5 & 2236 & +49.9 \\
        EquiRec & 994 & 1254 & +26.2 & 1048 & +5.4 & 1304 & +31.2 \\
        Recon & 1044 & 1482 & +42.0 & 1128 & +8.0 & 1506 & +44.3 \\
        FullRecon & 1094 & 1645 & +50.4 & 1161 & +6.1 & 1652 & +51.0 \\
        FullPoly & 1398 & 2086 & +49.2 & 1511 & +8.1 & 2019 & +44.4 \\
        FullOmega & 1451 & 2352 & +62.1 & 1582 & +9.0 & 2308 & +59.1 \\
      \hline
        Total & 21746 & 31024 & +42.7 & 23260 & +7.0 & 30795 & +41.6 \\
      \hline
        \multicolumn{8}{c}{}
    \end{tabular}
    \caption{Execution time of four implementations.}
    \label{tab:ext-comparison}
\end{table}

>>>>

\paragraph{Extracting Language Components}
Using the pattern demonstrated in Section~\ref{subsec:language-component}, we extract
reusable components from all the calculi. Each
component, which may contain several syntactical structures,
represents a certain feature of the language. For
example, the \lstinline{VarApp} component below represents variables and
function applications.

\lstinputlisting[linerange=13-29]{code/src/papercode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_VARAPP

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{resources/depGraph.pdf}
    \caption{Dependency graph of all calculi and components in case study.}
    \label{fig:dependency}
\end{figure*}

As shown in the graph, some components such as \lstinline{VarApp} are
created from scratch, while others such as \lstinline{Typed} are
extended from existing components. Since calculi and components have similar signatures, each calculus
can also be extended and reused directly. For example, calculus \lstinline{FullRef} extends from
calculus \lstinline{FullSimple}.

From the dependency graph, we know
that common components such as \lstinline{VarApp} are reused in
lots of calculi. Such reuse could shorten the code considerably. We
will show this advantage and inspect the possible performance penalty in
a quantitative way in the next subsection.

@@

\begin{table*}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \multicolumn{3}{ c| }{\bfseries SLOC} & \multicolumn{3}{ c| }{\bfseries Time (ms)} \\ \cline{2-7}
        \multicolumn{1}{|c|}{} & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% \\
      \hline
      \input{resources/CaseStudyTable.tex}
      \hline
      \multicolumn{7}{c}{}
    \end{tabular}
    \caption{Comparison of SLOC and execution time.}
    \label{tab:comparison}
\end{table*}

@@

\begin{table*}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \ilyaimpl{} & \multicolumn{2}{ c| }{\ourimpl{}} & \multicolumn{2}{ c| }{\ilyalongest{}} & \multicolumn{2}{ c| }{\ourclass{}} \\ \cline{2-8}
        \multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{\bfseries Time} & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% \\
      \hline
        Arith & 741 & 913 & +23.2 & 793 & +7.0 & 932 & +25.8 \\
        Untyped & 770 & 1018 & +32.2 & 821 & +6.6 & 1007 & +30.8 \\
        FullUntyped & 1297 & 1854 & +42.9 & 1343 & +3.5 & 1767 & +36.2 \\
        TyArith & 746 & 888 & +19.0 & 772 & +3.5 & 918 & +23.1 \\
        SimpleBool & 1376 & 1782 & +29.5 & 1494 & +8.6 & 1824 & +32.6 \\
        FullSimple & 1441 & 2270 & +57.5 & 1574 & +9.2 & 2226 & +54.5 \\
        Bot & 1080 & 1287 & +19.2 & 1078 & -0.2 & 1306 & +20.9 \\
      %\hline
        %\multicolumn{1}{|c|}{\dots} & \multicolumn{7}{c|}{\dots} \\
      %\hline
        FullRef & 1438 & 2291 & +59.3 & 1544 & +7.4 & 2142 & +49.0 \\
        FullError & 1410 & 1946 & +38.0 & 1524 & +8.1 & 1981 & +40.5 \\
        RcdSubBot & 1247 & 1524 & +22.2 & 1285 & +3.0 & 1612 & +29.3 \\
        FullSub & 1320 & 1979 & +49.9 & 1393 & +5.5 & 1899 & +43.9 \\
        FullEquiRec & 1407 & 2200 & +56.4 & 1561 & +10.9 & 2156 & +53.2 \\
        FullIsoRec & 1492 & 2253 & +51.0 & 1648 & +10.5 & 2236 & +49.9 \\
        EquiRec & 994 & 1254 & +26.2 & 1048 & +5.4 & 1304 & +31.2 \\
        Recon & 1044 & 1482 & +42.0 & 1128 & +8.0 & 1506 & +44.3 \\
        FullRecon & 1094 & 1645 & +50.4 & 1161 & +6.1 & 1652 & +51.0 \\
        FullPoly & 1398 & 2086 & +49.2 & 1511 & +8.1 & 2019 & +44.4 \\
        FullOmega & 1451 & 2352 & +62.1 & 1582 & +9.0 & 2308 & +59.1 \\
      \hline
        Total & 21746 & 31024 & +42.7 & 23260 & +7.0 & 30795 & +41.6 \\
      \hline
        \multicolumn{8}{c}{}
    \end{tabular}
    \caption{Execution time of four implementations.}
    \label{tab:ext-comparison}
\end{table*}

### 2017-04-27 SLE_17/sections/Discussion.tex
Some changes

We take the example in Section~\ref{subsec:parsingwithoa} again, which is a language of literals, additions and variables. It is defined as a
language component on the left side of Figure~\ref{fig:lng-components}.
Then for the extension of types and lambda abstractions in Section~\ref{subsec:differentsyntax}, instead of inheriting from the previous language directly, we define it as another independent language component. The code is presented on the right side of Figure~\ref{fig:lng-components}.

\begin{figure}[t]
\lstinputlisting[linerange=115-130]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
\lstinputlisting[linerange=134-150]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\caption{Language components \lstinline{VarExpr} (\huang{todo}left) and \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

As shown in Figure~\ref{fig:compose-components}, those two language components can be merged into one to obtain the language we want. Furthermore, the new language is still a modular
component ready for future composition. In that case modularity is realized over higher-order hierarchies.

\begin{figure}[t]
\lstinputlisting[linerange=154-165]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
\caption{Reuse two language components \lstinline{VarExpr} and \lstinline{TypedLam}.}\label{fig:compose-components}
\end{figure}

>>>>

We take the example in Section~\ref{subsec:parsingwithoa} again, which is a language of literals, additions and variables. It can be defined as a language component \inlinecode{VarExpr}.

\lstinputlisting[linerange=115-130]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR

For the extension of types and lambda abstractions in Section~\ref{subsec:differentsyntax}, instead of inheriting from the previous language directly, we can define it as another independent language component \inlinecode{TypedLam}.

\lstinputlisting[linerange=134-150]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM

The code below shows how we merge those two components together to obtain the language we want. Furthermore, the new language is still a modular
component ready for future composition. In that case modularity is realized over higher-order hierarchies.

\lstinputlisting[linerange=154-165]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR

### 2017-04-27 SLE_17/sections/Introduction.tex
Code & content
Some changes

So far research on semantic modularization techniques has focused on
operations that \emph{traverse} or \emph{process} extensible
data structures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce ASTs.  In particular the
problem of how to modularize parsing has not been studied in semantic
modularization approaches. At best, techniques such as NOA~\cite{Gouseti2014}
employ a syntactic modularity approach for parsing in combination with
a semantic modularity approach for defining other language components
that traverse or process ASTs. This is a shame because parsing is a
fundamental part of PL implementations, and it ought to be made
semantically modular as well, so that the full benefits of semantic
modularity apply.

@@

\begin{itemize}

>>>>

So far research on semantic modularization techniques has focused on
operations that \emph{traverse} or \emph{process} extensible
data structures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce ASTs.  In particular the
problem of how to modularize parsing has not been studied in semantic
modularization approaches. At best, techniques such as NOA~\cite{Gouseti2014}
employ a syntactic modularity approach for parsing in combination with
a semantic modularity approach for defining operations
that traverse or process ASTs. This is a shame because parsing is a
fundamental part of PL implementations, and it ought to be made
semantically modular as well, so that the full benefits of semantic
modularity apply.

@@

\begin{itemize}[leftmargin=*]

### 2017-04-27 SLE_17/sections/PackratParsing.tex
Some changes

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators~\cite{Ford2002}
are suitable to address those challenges. The algorithmic challenges
are important because they rule out various common techniques
used by non-modular code using parser combinators.
To avoid pitfalls related to those algorithmic challenges,
we propose the following methodology:

>>>>

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators~\cite{Ford2002}
are suitable to address them. The algorithmic challenges
are important because they rule out various common techniques
used by non-modular code using parser combinators.
To avoid pitfalls related to those algorithmic challenges,
we propose the following methodology:
